\pkg{cpop}: Detecting Changes in Piecewise-Linear Signals

### 2022-11-21 latex/response.tex

\title{cpop: Detecting changes in piecewise-linear signals -- response to review and pre-screening comments}
\author{Paul Fearnhead and Daniel Grose}
\date{\today}

\documentclass[12pt]{article}

\usepackage[dvipsnames]{xcolor}

\begin{document}
\maketitle

\noindent
{\bf Response to pre-screening comments}
\\

Thanks for the careful reading of our paper, and the helpful comments. We apologise for not fully meeting the journal's stylistic requirements. We have now updated the paper as requested:
\begin{itemize}
\item[(i)] We have replaced the use of cat() with warning() in the code.
\item[(ii)] We have used the third (rather than first) argument of system.time when calculating the plot comparing timings of methods.
\item[(iii)] We have corrected the incorrect code chunks for loops etc. (i.e., replacing R $>$ with +).
\item[(iv)] The section and subsection titles are now in sentence style; citations are in title style.
\item[(v)] We have added commas after i.e. and e.g. and ensured we have used prog, code and proglang throughout.
\item[(vi)] Where they have citation info, we have ensured that all R packages are cited per this information.
\end{itemize}

For convenient we append below the summary and point-by-point response to the reviewers' comments on the previous version of the paper.

\pagebreak
\noindent
{\bf Response to reviewers' comments}
\\

Below we reproduce the comments on the original version of our paper ``cpop: Detecting changes in piecewise-linear signals", together with a point-by-point responses and summary of the changes we have made to the paper and the package. The original comments are in italics.

\begin{itemize}
\item {\em The current package version is 1.0.3. Unfortunately the package does not contain a NEWS file which would allow to easily assess which functionality was added since version 0.0.3. Presumably the extensions cover unevenly spaced data, heterogeneous noise variance, and the possibility of a grid of potential change locations different from the locations of the data points.}

We have now added a NEWS file.

\item {\em The manuscript contains a section on R packages for changepoint detection. However, this section fails to relate the functionality provided by these packages to the functionality provided by the presented package. It would seem important to highlight for which use cases which packages would seem appropriate to guide potential users. In addition for the Journal of Statistical Software the software overview should cover statistical software packages in general, not only R packages.}

We have changed the title of this section. However, we are unaware of many software packages outside of R that perform changepoint detection, and none that perform change-in-slope. There is one package in Python, called ruptures, that has only been developed recently and claims to be the first Python package for changepoints. However this does not do change-in-slope. We have added a reference to this package within the paper.

The paper emphasises the difference in the methods for change-in-slope detecting in the first part of the Introduction (see also Figure 1), and now states more clearly the extra functionality of the cpop package over the packages for NOT and trend-filtering.

\item {\em Function cpop has an argument sd with default value 1. It seems questionable that this is a reasonable default value for any kind of data but maybe rather a data-driven default value would seem more appropriate. To enhance the usability of the package more guidance should be provided on how this argument should be chosen and for example provide a function argument specification which determines a data-driven value. Note that the Journal of Statistical Software aims at publishing statistical software which particularly helps users in their data analysis providing suitable tools to do not conveniently.}

We have now implemented a default method for estimating the standard deviation of the noise -- based on the variance of the second differences of the data. We have also added a warning if the user does not specify the standard deviation (and thus this default is used). We have similarly added a warning if the user does not specify the penalty (beta) -- as errors in specifying sd or beta have the same affect of impacting the estimate of the number of changepoints.

How to estimate the noise standard deviation is covered in some of the examples int he paper, and in particular we suggest using the CROPS function to run CPOP with a range of beta values (which is equivalent to running it for a fixed beta value and a range of sd values).

\item {\em Loading the package gives:}
\begin{verbatim}
> library("cpop")
Loading required package: crops
Loading required package: pacman
Attaching package: ‘cpop’
The following object is masked from ‘package:stats’:
fitted
The following object is masked from ‘package:methods’:
show
\end{verbatim}
{\em Clearly the masking should be avoided. E.g., by including}
\begin{verbatim}
    setGeneric("fitted", package = "stats")
\end{verbatim}
{\em and also importing show from methods.}

The methods shown and fitted in the cpop package were being set as generic methods with no reference to any existing S4 method of that name. This has now been prevented by including a test to see if these methods have already been set as generic, for instance, by another package.

\item {\em The output of}
\begin{verbatim}
> help(package = "cpop")
\end{verbatim}
{\em indicates that the vignette name "article" might not be very suitable and the documentation}
\begin{verbatim}
cpop cpop
\end{verbatim}
{\em is not very informative. Also the titles are not consistent in title or sentence style.
The description contains (2018) but it seems the paper was published in 2019.}

The vignette has been renamed to cpop. The documentation for the cpop function has been extended and the titles
for all sections of the documentation has been changed to sentence style. The date has been corrected in the packages DESCRIPTION file.

\item {\em The code defines twice a function null2na which is not called within the package and also not exported.}

The redundant functions have been removed from the package.

\item {\em Running example(cpop) and then using estimate() gives the following results:}
\begin{verbatim}
> estimate(res.true, x = 100)
x y_hat
1 100 11.0609
> estimate(res.true, x = res.true@x[50])
x y_hat
100 11.0609
> tail(estimate(res.true))
x y_hat
45 81.00 10.37956
46 84.64 10.43842
47 88.36 10.49856
48 92.16 10.56001
49 96.04 10.62274
50 100.00 10.68677
\end{verbatim}
{\em It is unclear why results differ for specific locations in case more locations are simultaneously evaluated.}

This incorrect behaviour was a consequence of a coding error in the estimate funciton. This has now been corrected.

\item {\em The code shown in the manuscript is}
\begin{verbatim}
R> res <- cpop(y, x, grid, sd = sqrt(sig2), minseg = 0.2, 
beta = 2*log(200))    
\end{verbatim}
{\em with comment in the text:
To avoid the potential for adding multiple changepoints between two observations, we set a minimum segment length of 0.09 (as the largest distance between consecutive x values is 0.08).
The code in the replication material then reads:}
\begin{verbatim}
res <- cpop(y, x, grid, sd=sqrt(sig2), minseg = 0.09,
beta=2*log(200))
\end{verbatim}

The code in the manuscript has been changed 
so that the article and the reproduction script are now consistent.

\end{itemize}
\end{document}

### 2022-11-21 latex/article.tex
Add files via upload

\subsection{Software for Changepoint Detection}

Currently most software for changepoint detection is available in \proglang{R} \cite[the main exception being the \pkg{ruptures} \proglang{Python} package of][which has similar functionality to the \pkg{changepoint } described below]{truong2018ruptures}.
There are both many different types of change that one may wish to detect, and many different approaches to detecting multiple changes. Consequently there are a wide range of change algorithms with associated packages in \proglang{R}. For example the \pkg{changepoint} package \citep{killick2014changepoint} implements dynamic programming algorithms, such as PELT, for detecting changes in mean, variance or mean and variance. Other dynamic programming algorithms include \pkg{fpop} and \pkg{gfpop} \citep{runge2020gfpop} that implements the functional optimal partitioning algorithm \citep{maidstone2017optimal} for detecting changes in mean, with the latter package allowing for flexibility as to how the mean changes (such as monotonically increasing) and for different loss functions for measuring fit to data. The \pkg{breakfast} package implements a range of methods based on recursively applying a test to detect a single change, for example wild binary segmentation \citep{fryzlewicz2014wild} and IsolateDetect \citep{anastasiou2022detecting}; whilst \pkg{mosum} \citep{meier2021mosum} implements the MOSUM procedure. Packages \pkg{stepR} and \pkg{FDRseg} implement the multiscale approaches of \cite{frick2014multiscale}, \cite{pein2017heterogeneous} and \cite{li2016fdr}.

@@

Our interest is in estimating the number of changes in slope, $K$, their locations, $\tau_1,\ldots,\tau_K$, and the underlying signal. The latter is equivalent to estimating $f(\tau_i)$ for $i=0,\ldots,K+1$. To simplify notation we will denote these values by $\alpha_0,\ldots,\alpha_{K+1}$, i.e.~ $\alpha_i=f(\tau_i)$ for $i=0,\ldots,K+1$. Also, for this and other quantities we will use the shorthand $\alpha_{i:j}$ for integers $i\leq j$ to be the ordered set of values, $\alpha_i,\ldots,\alpha_j$.

\subsection{An $L_0$-Penalised Criteria}

@@

The choice of $\beta$ in (\ref{eq:penalised_cost}) is important for accurate estimates, with lower values of $\beta$ leading to larger estimates of $K$, the number of changes. If the noise is approximately Gaussian and independent, and the estimate of the noise variance is good, then $\beta=2\log n$ is an appropriate choice  \citep{fearnhead2019detecting}. In general these assumptions will not hold, and often larger values of $\beta$ are required to e.g. compensate for positively auto-correlated noise. Where possible we recommend evaluating sets of estimated changepoints obtained for a range of $\beta$ values, and these can obtained in a computationally efficient manner using the CROPS algorithm of \cite{haynes2017computationally}.

\subsection{Dynamic Programming Recursion}

Solving (\ref{eq:penalised_cost}) is non-trivial as it involves minimising a non-convex function. Furthermore, standard dynamic programming algorithms for change points \citep{maidstone2017optimal}, e.g. those that recurse based on conditioning on the location of the most recent changepoint,  cannot be used because of the dependence across changepoints due to the continuity constraint. Thus we follow \cite{fearnhead2019detecting} and develop a dynamic programming recursion that conditions both on the location of the changepoints and the value of the mean at that changepoint.

Remember that we are allowing changepoints only at grid-points, $g_{1:N}$. We introduce a set of functions, each associated with a grid-point, $F_t(\alpha)$ for $t=1,\ldots,N$, defined as the minimum value of the penalised cost for data up to and including grid-point $g_t$ conditional on the signal at $g_t$ being $\alpha$, i.e. $f(g_t)=\alpha$. To define this formally, define a set of segment cost functions
\[
\mathcal{C}_{s,t}(\alpha',\alpha) = \sum_{i=n_s+1}^{n_t} \left(y_i -\alpha' - (\alpha-\alpha')\frac{x_i-g_s}{g_t-g_s} \right)^2,
\]
which is the cost of fitting a linear signal to data points between the $s$th and $t$th grid-points, i.e. $(x_i,y_i)$ with $g_s<x_i\leq g_t$, with the signal taking the value $\alpha'$ at $g_s$ and $\alpha$ at $g_t$. In the summation we use the notation $n_s$ to denote the index of the last data-point located at or before $g_s$. If $n_t=n_s$, that is there are no data-points between $g_s$ and $g_t$ then this cost is set to 0.

@@

\section{The {cpop} Package} \label{sec:cpop}

@@

\subsection{Generating Simulated Data}
%
%
The \code{simchangeslope} function allows for simulating data from a change-in-slope model (\ref{eq:cinslope}): %. It takes as input the $x$-values of the data points, the locations of the changes, how much the slop of the mean changes at each changepoint, and the standard deviation of the noise:
\begin{CodeInput}
simchangeslope(x, changepoints, change.slope, sd = 1)
\end{CodeInput}
%
%
It takes the following arguments :
\begin{itemize}
\item \code{x} -  A numeric vector containing the locations of the data.
\item \code{changepoints} - A numeric vector of changepoint locations.
\item \code{change.slope} -  A numeric vector indicating the change in slope at each changepoint. The initial slope is assumed to be 0.
\item \code{sd} - The residual standard deviation. Can be a single numerical value or a vector of values for the case of varying residual standard deviation. Default value is 1.
\end{itemize}
%
%
It returns a vector $y$ of simulated values which correspond to the locations $x$. 
%
%
The mean function of the data goes through the origin -- but to add an intercept we just add a constant to all output values. It is possible to get the value of the mean function at the $x$-values of the data by setting \code{sd=0}.

@@

\subsection{Determining Changes in Slope}

@@

\item \code{minseglen} -  The minimum allowable segment length, i.e. distance between successive changepoints. The default is that no minimum segment length is imposed.

@@

\begin{figure}
\centering
%\includegraphics[scale=1.0]{figures/cpop-example1.pdf}
\includegraphics[width=0.6\linewidth]{figures/cpop_example1_ggplot.pdf}
\caption{Example output of \code{cpop} for simulated data from Figure \ref{fig:simulate-example}. The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines).   }
\label{fig:cpop-example}
\end{figure}
%
%
\subsection{Other Functions}
%
%
In addition to \code{plot} and \code{summary}, the \pkg{cpop} package provides functions to evaluate the fitted mean function at specified $x$-values, and to calculate the residuals of the fitted mean. The primary argument of these functions is \code{object}, an instance of a cpop S4 class as produced by the function \code{cpop}.

@@

\section{Extensions of {cpop}} \label{sec:extensions}

\subsection{Irregularly Sampled Data}

@@

\begin{figure}
\centering
%\includegraphics{figures/cpop-example-uneven.pdf}
\includegraphics[width=0.6\linewidth]{figures/cpop_example_uneven_ggplot.pdf}
\caption{Example output of \code{cpop} for unevenly spaced simulated data. The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines). }
\label{fig:cpop-example-uneven}
\end{figure}
%
%
\subsection{Heterogeneous Data}

@@

\subsection{Choice of Grid}

@@

In our example we find 38 changepoints with this coarse grid. We then introduce a finer grid around these putative changes: our new grid includes all $x$-values within 8 of each putative changepoint.
\begin{CodeChunk}
\begin{CodeInput}
R> cps <- unlist(changepoints(res.coarse))

R> grid <- NULL
R> for(i in 1:length(cps))
R> {
R>   grid <- c(grid, cps[i]+(-7):8)
R> }

R> res.fine <- cpop(y, x, grid, beta = 2*log(length(x)), sd = 1)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Imposing a Minimum Segment Length}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> changepoints <- c(0, 25, 50, 100)
R> change.slope <- c(0.2, -0.3, 0.2, -0.1)
R> x <- 1:200
R> mu <- simchangeslope(x, changepoints, change.slope, sigma = 0)

R> y <- mu + rt(length(x), df = 4)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Choice of Penalty}

@@

We can plot the location in the changepoints for each segmentation found by CPOP for $\beta\in[5,50]$
\begin{CodeInput}
R> plot(res.crops)
\end{CodeInput}
This is shown in Figure \ref{fig:crops.cpopA}, and shows that there are 6 different segmentations found. These are labelled with a penalty value which gives that segmentation (left axis) and the unpenalised cost, i.e. the weighted RSS, for that segmentation and penalty value (right axis).

Details of the segmentations can be obtained using \code{segmentations(res.crops) }. This gives a matrix with one row for each of the segmentations, and each row contains a value of $\beta$ that gives that segmentation, the corresponding unpenalised cost, the penalised cost, the number of changepoints, and then the list of ordered changepoints. 
%
%
We can also obtain a list with the output from \code{cpop} corresponding to each segmentation, with \code{models( res.crops )}. For example, one approach to choose a segmentation from the output from \code{cpop.crops} is to find the segmentation that minimises a cost under a model where we assume the noise variance is unknown \citep{fryzlewicz2014wild},
\[
 n \log \left( \frac{1}{n} \sum_{i=1}^n \left(y_i-\hat{f}(x_i)\right)^2
\right) + 2 K \log n.
\]
Here $\hat{f}$ is the estimated mean function and $K$ is the number of changepoints. This can be calculated as follows.
\begin{CodeChunk}
\begin{CodeInput}
R> models <- cpop.crops.models(res.crops)
R> M <- length(models)
R> BIC <- rep(NA, M)
R> ncps <- segmentations(res.crops )[,4]
R> n <- length(y)
R> for(j in 1:M)
R> {
R>   BIC[i] <- n*log(mean((residuals(models[[j]]))^2)) + 2*ncps[i]*log(n)
R> }
\end{CodeInput}
\end{CodeChunk}
This uses that the fourth column of the matrix \code{segmentations( res.crops )} stores the number of changepoints in each segmentation, and that we can calculate $y_i - \hat{f}(x_i)$ using the \code{residual} function evaluated for the corresponding entry of \code{cpop.crops.models(res.crops)}. The segmentation which has the smallest value of \code{BIC} is shown in Figure \ref{fig:crops.cpopA}, and shows that this correctly chooses the segmentation with three changes.

@@

Interest lies in estimating the rate of decay of the log-spectra against log-wavenumber. We can do this by removing the first three data points (where the spectra is increasing) and then using \code{cpop} to fit a piecewise-linear curve to the remaining data. We perform an initial run of \code{cpop} assuming an estimated homogeneous noise variance on the data from August from the 2000 run.
\begin{CodeChunk}
\begin{CodeInput}
R> data("wavenumber_spectra")
R> x <- log(wavenumber_spectra[-(1:3),1])
R> y <- log(wavenumber_spectra[-(1:3),4])

R> grid <- seq(from = min(x), to = max(x), length = 200)

R> sig2 <- mean(diff(diff(y))^2)/6

R> res <- cpop(y, x, grid, sd = sqrt(sig2), 
+              minseglen = 0.09, beta = 2*log(200))
\end{CodeInput}
\end{CodeChunk}
Here we estimate the noise variance, \code{sig2}, based on the variance of the double difference of the data. For regions where the mean is linear and the data is evenly spaced, taking the double difference will lead to a mean zero process. If the noise is IID with variance $\sigma^2$ then the double-differenced process will have a marginal variance that is $6\sigma^2$. Thus our estimate is the empirical mean of the square of the double-difference data divided by 6. The original data is evenly spaced in in terms of wavenumber, but as we take logs, \code{x} is unevenly spaced: so this estimator will be biased in our setting. However it will give a reasonable ball-park figure for an initial run of \code{cpop} -- the residuals from which can then be used to get a better estimate of the noise variance. We use a evenly spaced grid for possible change-point locations. To avoid the potential for adding multiple changepoints between two observations, we set a minimum segment length of 0.09 (as the largest distance between consecutive \code{x} values is 0.08).

@@

The output is shown in the top-right plot of Figure \ref{fig:crops.data}, and appears to be over-fitting to the early part of the series. This is because the noise variance is heterogeneous, and decreasing with \code{x}. However, given our initial fit we can use the residuals to estimate the noise variance. The noise for the spectra is expected to be approximately inversely proportional to the wavenumber. By using a Taylor-expansion, we have the variance of the noise for the log-spectra should be approximately the variance of the noise of the spectra divided by the square of the mean of the spectra. As the mean of the spectra is roughly a power of the wavenumber, this suggests using a model for the variance, $\sigma^2_x$ say, depending on $x$ as $\log \sigma^2_x = a + bx$: so that the variance is proportional to some power of the wavenumber. We can estimate the parameters of this model by maximising the log-likelihood of Gaussian model for the residuals with this form for the variance.
\begin{CodeChunk}
\begin{CodeInput}
R> r2 <- residuals(res)^2
R> loglik <- function(par)
R> { 
R>  return(length(r2)*par[1] + par[2]*sum(x) + 
+          sum(r2/(exp(par[1]+par[2]*x))))
R> }
R> est.hat <- optim(c(0,0), loglik)
R> sig2 <- exp(est.hat$par[1] + est.hat$par[2]*x)
R> res2 <- cpop(y, x, grid, sd = sqrt(sig2), 
+               minseglen = 0.09, beta = 2*log(200))
\end{CodeInput}
\end{CodeChunk}
Here we have calculated the maximum likelihood estimates by using \code{optim} to minimise minus the log-likelihood. The resulting output from \code{cpop} is shown in the bottom left plot of Figure \ref{fig:crops.data}.  The first two changes could represent real regime transitions relating to the inviscid fluid physics that one would see in the real ocean \cite[see Figure 6a of][]{callies2013interpreting}, while the three changes for the largest values of \code{x} may relate to a breakdown in the numerical ocean model near the highest wavenumber of the ocean model grid \cite[]{soufflet2016effective}. The estimates of the spectra for all four series, obtained by repeating this approach, is also shown in Figure \ref{fig:crops.data}. For this application, the residuals from the fitted model appear to be uncorrelated and sub-Gaussian, so using the fit based on the default penalty choice is reasonable. Though one could also explore segmentations for other penalty choices using \code{cpop.crops} as in the previous section.

@@

\section{Details of the Recursion} \label{App:cost_update}

>>>>

\subsection{Software for changepoint detection}

Currently most software for changepoint detection is available in \proglang{R} \cite[the main exception being the \pkg{ruptures} \proglang{Python} package of][which has similar functionality to the \pkg{changepoint } described below]{truong2018ruptures}.
There are both many different types of change that one may wish to detect, and many different approaches to detecting multiple changes. Consequently there are a wide range of change algorithms with associated packages in \proglang{R}. For example the \pkg{changepoint} package \citep{killick2014changepoint} implements dynamic programming algorithms, such as PELT, for detecting changes in mean, variance or mean and variance. Other dynamic programming algorithms include \pkg{fpop} and \pkg{gfpop} \citep{runge2020gfpop} that implements the functional optimal partitioning algorithm \citep{maidstone2017optimal} for detecting changes in mean, with the latter package allowing for flexibility as to how the mean changes (such as monotonically increasing) and for different loss functions for measuring fit to data. The \pkg{breakfast} package implements a range of methods based on recursively applying a test to detect a single change, for example wild binary segmentation \citep{fryzlewicz2014wild} and IsolateDetect \citep{anastasiou2022detecting}; whilst \pkg{mosum} \citep{meier2021mosum} implements the MOSUM procedure. Packages \pkg{stepR} \cite[]{stepR} and \pkg{FDRseg} implement the multiscale approaches of \cite{frick2014multiscale}, \cite{pein2017heterogeneous} and \cite{li2016fdr}.

@@

Our interest is in estimating the number of changes in slope, $K$, their locations, $\tau_1,\ldots,\tau_K$, and the underlying signal. The latter is equivalent to estimating $f(\tau_i)$ for $i=0,\ldots,K+1$. To simplify notation we will denote these values by $\alpha_0,\ldots,\alpha_{K+1}$, so $\alpha_i=f(\tau_i)$ for $i=0,\ldots,K+1$. Also, for this and other quantities we will use the shorthand $\alpha_{i:j}$ for integers $i\leq j$ to be the ordered set of values, $\alpha_i,\ldots,\alpha_j$.

\subsection{An $L_0$-penalised criteria}

@@

The choice of $\beta$ in (\ref{eq:penalised_cost}) is important for accurate estimates, with lower values of $\beta$ leading to larger estimates of $K$, the number of changes. If the noise is approximately Gaussian and independent, and the estimate of the noise variance is good, then $\beta=2\log n$ is an appropriate choice  \citep{fearnhead2019detecting}. In general these assumptions will not hold, and often larger values of $\beta$ are required to compensate for positively auto-correlated noise. Where possible we recommend evaluating sets of estimated changepoints obtained for a range of $\beta$ values, and these can obtained in a computationally efficient manner using the CROPS algorithm of \cite{haynes2017computationally}.

\subsection{Dynamic programming recursion}

Solving (\ref{eq:penalised_cost}) is non-trivial as it involves minimising a non-convex function. Furthermore, standard dynamic programming algorithms for change points \citep{maidstone2017optimal}, e.g., those that recurse based on conditioning on the location of the most recent changepoint,  cannot be used because of the dependence across changepoints due to the continuity constraint. Thus we follow \cite{fearnhead2019detecting} and develop a dynamic programming recursion that conditions both on the location of the changepoints and the value of the mean at that changepoint.

Remember that we are allowing changepoints only at grid-points, $g_{1:N}$. We introduce a set of functions, each associated with a grid-point, $F_t(\alpha)$ for $t=1,\ldots,N$, defined as the minimum value of the penalised cost for data up to and including grid-point $g_t$ conditional on the signal at $g_t$ being $\alpha$, i.e., $f(g_t)=\alpha$. To define this formally, define a set of segment cost functions
\[
\mathcal{C}_{s,t}(\alpha',\alpha) = \sum_{i=n_s+1}^{n_t} \left(y_i -\alpha' - (\alpha-\alpha')\frac{x_i-g_s}{g_t-g_s} \right)^2,
\]
which is the cost of fitting a linear signal to data points between the $s$th and $t$th grid-points, i.e., $(x_i,y_i)$ with $g_s<x_i\leq g_t$, with the signal taking the value $\alpha'$ at $g_s$ and $\alpha$ at $g_t$. In the summation we use the notation $n_s$ to denote the index of the last data-point located at or before $g_s$. If $n_t=n_s$, that is there are no data-points between $g_s$ and $g_t$ then this cost is set to 0.

@@

\section[The cpop package]{The \pkg{cpop} package} \label{sec:cpop}

@@

\subsection{Generating simulated data}
%
%
The \code{simchangeslope} function allows for simulating data from a change-in-slope model (\ref{eq:cinslope}): %. It takes as input the $x$-values of the data points, the locations of the changes, how much the slop of the mean changes at each changepoint, and the standard deviation of the noise:
\begin{CodeInput}
simchangeslope(x, changepoints, change.slope, sd = 1)
\end{CodeInput}
%
%
It takes the following arguments :
\begin{itemize}
\item \code{x} -  A numeric vector containing the locations of the data.
\item \code{changepoints} - A numeric vector of changepoint locations.
\item \code{change.slope} -  A numeric vector indicating the change in slope at each changepoint. The initial slope is assumed to be 0.
\item \code{sd} - The residual standard deviation. Can be a single numerical value or a vector of values for the case of varying residual standard deviation. Default value is 1.
\end{itemize}
%
%
It returns a vector $y$ of simulated values which correspond to the locations $x$. 
%
%
The mean function of the data goes through the origin -- but to add an intercept we just add a constant to all output values. It is possible to get the value of the mean function at the $x$-values of the data by setting \code{sd=0}.

@@

\subsection{Determining changes in slope}

@@

\item \code{minseglen} -  The minimum allowable segment length, that is the distance between successive changepoints. The default is that no minimum segment length is imposed.

@@

\begin{figure}
\centering
%\includegraphics[scale=1.0]{figures/cpop-example1.pdf}
\includegraphics[width=0.6\linewidth]{figures/cpop_example1_ggplot.pdf}
\caption{Example output of \code{cpop} for simulated data from Figure \ref{fig:simulate-example}. The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines).   }
\label{fig:cpop-example}
\end{figure}
%
%
\subsection{Other functions}
%
%
In addition to \code{plot} and \code{summary}, the \pkg{cpop} package provides functions to evaluate the fitted mean function at specified $x$-values, and to calculate the residuals of the fitted mean. The primary argument of these functions is \code{object}, an instance of a cpop S4 class as produced by the function \code{cpop}.

@@

\section[Extensions of {cpop}]{Extensions of \pkg{cpop}} \label{sec:extensions}

\subsection{Irregularly sampled data}

@@

\begin{figure}
\centering
%\includegraphics{figures/cpop-example-uneven.pdf}
\includegraphics[width=0.6\linewidth]{figures/cpop_example_uneven_ggplot.pdf}
\caption{Example output of \code{cpop} for unevenly spaced simulated data. The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines). }
\label{fig:cpop-example-uneven}
\end{figure}
%
%
\subsection{Heterogeneous data}

@@

\subsection{Choice of grid}

@@

In our example we find 38 changepoints with this coarse grid. We then introduce a finer grid around these putative changes: our new grid includes all $x$-values within 8 of each putative changepoint.
\begin{CodeChunk}
\begin{CodeInput}
R> cps <- unlist(changepoints(res.coarse))
R> grid <- NULL
R> for(i in 1:length(cps)) {
+   grid <- c(grid, cps[i]+(-7):8)
+ }
R> res.fine <- cpop(y, x, grid, beta = 2*log(length(x)), sd = 1)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Imposing a minimum segment length}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> changepoints <- c(0, 25, 50, 100)
R> change.slope <- c(0.2, -0.3, 0.2, -0.1)
R> x <- 1:200
R> mu <- simchangeslope(x, changepoints, change.slope, sigma = 0)
R> y <- mu + rt(length(x), df = 4)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Choice of penalty}

@@

We can plot the location in the changepoints for each segmentation found by CPOP for $\beta\in[5,50]$
\begin{CodeInput}
R> plot(res.crops)
\end{CodeInput}
This is shown in Figure \ref{fig:crops.cpopA}, and shows that there are 6 different segmentations found. These are labelled with a penalty value which gives that segmentation (left axis) and the unpenalised cost, i.e., the weighted RSS, for that segmentation and penalty value (right axis).

Details of the segmentations can be obtained using \code{segmentations(res.crops) }. This gives a matrix with one row for each of the segmentations, and each row contains a value of $\beta$ that gives that segmentation, the corresponding unpenalised cost, the penalised cost, the number of changepoints, and then the list of ordered changepoints. 
%
%
We can also obtain a list with the output from \code{cpop} corresponding to each segmentation, with \code{models( res.crops )}. For example, one approach to choose a segmentation from the output from \code{cpop.crops} is to find the segmentation that minimises a cost under a model where we assume the noise variance is unknown \citep{fryzlewicz2014wild},
\[
 n \log \left( \frac{1}{n} \sum_{i=1}^n \left(y_i-\hat{f}(x_i)\right)^2
\right) + 2 K \log n.
\]
Here $\hat{f}$ is the estimated mean function and $K$ is the number of changepoints. This can be calculated as follows.
\begin{CodeChunk}
\begin{CodeInput}
R> models <- cpop.crops.models(res.crops)
R> M <- length(models)
R> BIC <- rep(NA, M)
R> ncps <- segmentations(res.crops )[,4]
R> n <- length(y)
R> for(j in 1:M) {
+   BIC[i] <- n*log(mean((residuals(models[[j]]))^2)) + 2*ncps[i]*log(n)
+ }
\end{CodeInput}
\end{CodeChunk}
This uses that the fourth column of the matrix \code{segmentations( res.crops )} stores the number of changepoints in each segmentation, and that we can calculate $y_i - \hat{f}(x_i)$ using the \code{residual} function evaluated for the corresponding entry of \code{cpop.crops.models(res.crops)}. The segmentation which has the smallest value of \code{BIC} is shown in Figure \ref{fig:crops.cpopA}, and shows that this correctly chooses the segmentation with three changes.

@@

Interest lies in estimating the rate of decay of the log-spectra against log-wavenumber. We can do this by removing the first three data points (where the spectra is increasing) and then using \code{cpop} to fit a piecewise-linear curve to the remaining data. We perform an initial run of \code{cpop} assuming an estimated homogeneous noise variance on the data from August from the 2000 run.
\begin{CodeChunk}
\begin{CodeInput}
R> data("wavenumber_spectra")
R> x <- log(wavenumber_spectra[-(1:3),1])
R> y <- log(wavenumber_spectra[-(1:3),4])
R> grid <- seq(from = min(x), to = max(x), length = 200)

R> sig2 <- mean(diff(diff(y))^2)/6 
R> res <- cpop(y, x, grid, sd = sqrt(sig2), 
+              minseglen = 0.09, beta = 2*log(200))
\end{CodeInput}
\end{CodeChunk}
Here we estimate the noise variance, \code{sig2}, based on the variance of the double difference of the data. For regions where the mean is linear and the data is evenly spaced, taking the double difference will lead to a mean zero process. If the noise is IID with variance $\sigma^2$ then the double-differenced process will have a marginal variance that is $6\sigma^2$. Thus our estimate is the empirical mean of the square of the double-difference data divided by 6. The original data is evenly spaced in in terms of wavenumber, but as we take logs, \code{x} is unevenly spaced: so this estimator will be biased in our setting. However it will give a reasonable ball-park figure for an initial run of \code{cpop} -- the residuals from which can then be used to get a better estimate of the noise variance. We use a evenly spaced grid for possible change-point locations. To avoid the potential for adding multiple changepoints between two observations, we set a minimum segment length of 0.09 (as the largest distance between consecutive \code{x} values is 0.08).

@@

The output is shown in the top-right plot of Figure \ref{fig:crops.data}, and appears to be over-fitting to the early part of the series. This is because the noise variance is heterogeneous, and decreasing with \code{x}. However, given our initial fit we can use the residuals to estimate the noise variance. The noise for the spectra is expected to be approximately inversely proportional to the wavenumber. By using a Taylor-expansion, we have the variance of the noise for the log-spectra should be approximately the variance of the noise of the spectra divided by the square of the mean of the spectra. As the mean of the spectra is roughly a power of the wavenumber, this suggests using a model for the variance, $\sigma^2_x$ say, depending on $x$ as $\log \sigma^2_x = a + bx$: so that the variance is proportional to some power of the wavenumber. We can estimate the parameters of this model by maximising the log-likelihood of Gaussian model for the residuals with this form for the variance.
\newpage
\begin{CodeChunk}
\begin{CodeInput}
R> r2 <- residuals(res)^2
R> loglik <- function(par) { 
+  return(length(r2)*par[1] + par[2]*sum(x) + 
+          sum(r2/(exp(par[1]+par[2]*x))))
+ }
R> est.hat <- optim(c(0,0), loglik)
R> sig2 <- exp(est.hat$par[1] + est.hat$par[2]*x)
R> res2 <- cpop(y, x, grid, sd = sqrt(sig2), 
+               minseglen = 0.09, beta = 2*log(200))
\end{CodeInput}
\end{CodeChunk}
Here we have calculated the maximum likelihood estimates by using \code{optim} to minimise minus the log-likelihood. The resulting output from \code{cpop} is shown in the bottom left plot of Figure \ref{fig:crops.data}.  The first two changes could represent real regime transitions relating to the inviscid fluid physics that one would see in the real ocean \cite[see Figure 6a of][]{callies2013interpreting}, while the three changes for the largest values of \code{x} may relate to a breakdown in the numerical ocean model near the highest wavenumber of the ocean model grid \cite[]{soufflet2016effective}. The estimates of the spectra for all four series, obtained by repeating this approach, is also shown in Figure \ref{fig:crops.data}. For this application, the residuals from the fitted model appear to be uncorrelated and sub-Gaussian, so using the fit based on the default penalty choice is reasonable. Though one could also explore segmentations for other penalty choices using \code{cpop.crops} as in the previous section.

@@

\section{Details of the recursion} \label{App:cost_update}
